One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-10-26 doc/genesetdp.tex
Back to old name.

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{A simple null model for inferences from network enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Gustavo Jeuken\textsuperscript{1} and
Lukas K\"{a}ll\textsuperscript{1,*}
\\
\bigskip
\textbf{1} Science for Life Laboratory, School of Engineering Sciences in Chemistry, Biotechnology and Health, KTH -- Royal Institute of Technology, Box 1031, 17121 Solna, Sweden
\\
\bigskip

@@

Over-Representation Analysis (ORA) is commonly used to infer function from sets of analytes such as genes, transcripts, proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten,goeman2007analyzing}. The technique estimates which functional modules, such as complexes or pathways, are overrepresented among a set of identified analytes. One prominent application of the technique is expression analysis, where ORA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.

Most ORA methods are assessing the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test. However, variants such as Gene Set Enrichment Analysis (GSEA)\cite{subramanian2005gene} also includes information on expression levels of the analytes of the query set.

@@

The significance of the inferences from NEA is assessed by null models that investigate the number of expected random links from the query gene set to the pathway. Previous efforts have settled for network randomization, where the network definitions are randomly constructed by methods with varying degrees of topological conservation in respect to the original network.

Yet, the methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.

As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative.

@@

We argue that the network perturbation methods used in most flavours of NEA are missleading. Users of NEA do not differ from users of ORA and want to test an association between a query set of genes and pathways.  The typical user has used an experimental or computational method to derive a set of query genes and wants to know if it is a random set, or if it is enriched with respect to a pathway.  In this question, the introduction of a network is not relevant. A network may help to increase the sensitivity of ORA methods, but for practical reasons it is better left out of the null hypothesis. We argue that tests of significance in this type of analysis is naturally implemented as tests of the query set. That is, is the query set a randomly selected set of genes or is there an over-representation of well-connected genes in the query set?

Hence, here we employ a different null model, more directly aimed at modelling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. We present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC. Both algorithms circumvent the need of network pertubations for the calculations of significance.
Combining the GeneSetDP and GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{A simple null model for inferences from network enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Gustavo S. Jeuken\textsuperscript{1} and
Lukas K\"{a}ll\textsuperscript{1,*}
\\
\bigskip
\textbf{1} Science for Life Laboratory, School of Engineering Sciences in Chemistry, Biotechnology and Health, KTH -- Royal Institute of Technology, Box 1031, 17121 Solna, Sweden
\\
\bigskip

@@

Over-Representation Analysis (ORA) is commonly used to infer function from sets of analytes such as genes, transcripts, proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten,goeman2007analyzing}. The technique estimates which functional modules, such as complexes or pathways, are overrepresented among a set of identified analytes. One prominent application of the technique is expression analysis, where ORA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group. Most ORA methods are assessing the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test. However, variants such as Gene Set Enrichment Analysis (GSEA)\cite{subramanian2005gene} also includes information on expression levels of the analytes of the query set.

@@

The significance of the inferences from NEA is assessed by null models that investigate the number of expected random links from the query gene set to the pathway. Previous efforts have settled for network randomization, where the network definitions are randomly constructed by methods with varying degrees of topological conservation in respect to the original network. Yet, the methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints. As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments. While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative.

@@

We argue that the network perturbation methods used in most flavours of NEA are missleading. Users of NEA do not differ from users of ORA, they want to test an association between a query set of genes and pathways.  The typical user has used an experimental or computational method to derive a set of query genes and wants to know if it is a random set, or if it is enriched with respect to a pathway.  In this question, the introduction of a network is not relevant. A network may help to increase the sensitivity of ORA methods, but for practical reasons it is better left out of the null hypothesis. Tests of significance in this type of analysis is naturally implemented as tests of the query set. That is, is the query set a randomly selected set of genes or is there an over-representation of well-connected genes in the query set?

Hence, here we employ a different null model, more directly aimed at modelling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. Here, we present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC. Both algorithms circumvent the need of network pertubations for the calculations of significance. Combining the GeneSetDP and GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.
